---
title: "The C API"
---

```{r, include = FALSE}
# Every C example here is a file of the fixture package tools/zubintest,
# which CI builds, checks and tests on three operating systems; nothing is
# written for this page alone.
fixture <- function(path, lang = "c") {
  lines <- readLines(file.path("..", "..", "tools", "zubintest", path))
  cat("```", lang, "\n", paste(lines, collapse = "\n"), "\n```\n", sep = "")
}
```

zubin's R functions are built on a header-only C99 library, and any package can use the same
code: the buffer, the cursor, the typed reads and writes, and the record layouts with their
unpack and pack kernels. This page is for the author of such a package.

## Linking

There is nothing to link. Every function is `static inline` in a header, so a package that
includes them carries its own copy of what it uses, and zubin need not be installed when it
runs. The whole recipe is three files:

```{r, echo = FALSE, results = "asis"}
cat("`DESCRIPTION` (an excerpt):\n\n")
cat("```\n", grep("^LinkingTo", readLines("../../tools/zubintest/DESCRIPTION"), value = TRUE), "\n```\n", sep = "")
cat("\n`NAMESPACE`:\n\n")
fixture("NAMESPACE", "")
cat("\n`src/Makevars`:\n\n")
fixture("src/Makevars", "make")
```

`LinkingTo` names both zubin and zufast because it is not transitive: R adds each named
package's `include/` directory to the compiler's path, and zubin's headers include zufast's
`<zufast/bits.h>` and `<zufast/utf8.h>`. `$(C_VISIBILITY)` hides your symbols; the headers
emit nothing that is not already `static`, so two packages that both use zubin never bind to
each other's copies.

The contract is source compatibility, as for zufast: within major version 1 a function, type,
macro or enumerator that exists keeps existing with the same meaning, enumerator values are
permanent, and struct fields are only ever appended. Initialise every struct through its
`zb_*` function, never by aggregate initialisation, so an appended field has a defined value.
`ZUBIN_VERSION_NUMBER` is `10000 * major + 100 * minor + patch`.

## The headers

| Header | What it holds |
|---|---|
| `<zubin.h>` | everything below except `zubin-r.h` |
| `<zubin/status.h>` | `zb_status` and `zb_status_string()` |
| `<zubin/rw.h>` | `zb_rd_<type><order>()` and `zb_wr_...()` for every type at any alignment |
| `<zubin/cursor.h>` | `zb_cur`: checked sequential reads |
| `<zubin/buf.h>` | `zb_buf`: a buffer with ownership, a hard cap and checked growth |
| `<zubin/layout.h>` | record layouts, the spec parser, and the unpack and pack kernels |
| `<zubin-r.h>` | a `zb_buf` owned by an R external pointer; the only header that includes R's |

Every function that can fail returns a `zb_status`, which is `ZB_OK` (0) on success; results
come back through out-parameters. **On failure the inputs are unchanged**: a cursor that
returns `ZB_ERR_EOF` has not moved, a buffer that returns `ZB_ERR_LIMIT` holds what it held.
Nothing in the headers prints, aborts, exits or allocates, except `buf.h`, and that only
through `malloc`, `realloc` and `free`.

## Reading: the cursor

A cursor is a pointer, a length and a 0-based position, so every error a reader reports is
"at byte `c.pos`". One read per type and byte order: `zb_cur_u32le()`, `zb_cur_f64be()`,
`zb_cur_bytes()` to borrow a run of bytes, `zb_cur_seek()` and `zb_cur_skip()`.

```{r, echo = FALSE, results = "asis"}
fixture("src/cursor.c")
```

## Records: layouts and kernels

A layout is parsed from the same specification string `bin_layout()` takes, into a field
array the caller supplies, so the parser allocates nothing. Size the array with
`zb_layout_count_fields()`; names in the result are borrowed pointers into the
specification, which must outlive the layout.

The kernels then process one field of `n` records in one strided loop, into one typed column:
`zb_unpack_i32()` for the small integer types and `bool`, `zb_unpack_f64()` for `u32` and the
floating types, `zb_unpack_i64()` and `zb_unpack_f64x()` (exact or an error) for the 64-bit
integers, and `zb_unpack_bytes()` for `b` and `s`. Array fields are written column-major,
element `k` of record `i` at `dst[k * n + i]`. The `zb_pack_*()` kernels are the inverses, and
refuse a value that does not fit rather than wrapping it. The kernels do not bounds-check:
prove `offset + (n - 1) * stride + size <= length` first.

```{r, echo = FALSE, results = "asis"}
fixture("src/layout.c")
```

## Writing: the buffer, and buffers owned by R

A `zb_buf` is a pointer, a length, a capacity, a hard cap `max` (0 for none), flags, and a
`release` function with the `owner` it frees: the shape of the Arrow C data interface.
`zb_buf_alloc()` makes a growable buffer, `zb_buf_borrow()` a window onto your own memory
(writable up to its length, never grown), and `zb_buf_release()` frees whatever needs
freeing. Growth doubles to 64 MiB and grows by half after, never past `max`: an append that
would is `ZB_ERR_LIMIT`, sets `ZB_BUF_HIT_LIMIT`, and changes nothing. There is no bare size
arithmetic: sums and products that would overflow are `ZB_ERR_MEMORY`.

Ownership is a flag, never the identity of `release`: every translation unit has its own
copy of a header-only function, so comparing function pointers across units is meaningless.

In an R package, heap memory that must survive a longjmp from `Rf_error()` or
`R_CheckUserInterrupt()` must be owned by R before the first call that can jump.
`zb_r_buf_new()` creates the external pointer and registers its finalizer before the buffer
exists; `zb_r_buf_free()` releases it early, and the finalizer then does nothing:

```{r, echo = FALSE, results = "asis"}
fixture("src/buffer.c")
```

`zb_r_buf_get()` allocates nothing, so nothing needs protecting around it, and returns
`NULL` for an external pointer that is not one of these buffers or has been freed (one
restored from a saved session, say).
